Need Fulfillment During Intergroup Contact: Three Experience Sampling Studies

### 2023-05-26 manuscript.tex
wording

% Aim / Solution v.02: look at need fulfillment as a mechanism in daily interactions of migrants
% Proposal Paragraph:
We propose that one key to understanding how an interaction is perceived is to examine the level of need fulfillment it provided to an individual. As an example, if someone seeks acceptance by their interaction partner, and this need is fulfilled during the interaction, the person should rate the interaction and the group of the interaction partner more favorably. To test this proposal, we collected three sets of real-life data from recent immigrants, assessing their daily interactions with majority group members, tracking situational needs, interaction quality, well-being, and outgroup attitudes.

\section{Need Mechanism in Intergroup Contact}
% Motivational mechanism:
% interaction quality is important
% past research has either focused on conditions or cognitive-affective processes
% neither are necessary nor explain why an interaction is positive
\sout{Several meta-analytic reviews have found that positive intergroup contact reduces prejudice and increases positive attitudes in experimental and cross-sectional studies \citep[][]{Tropp2005, Pettigrew2006}, as well as in intergroup contact interventions outside the lab \citep[][]{Beelmann2014, Lemmer2015}.} When it comes to what constitutes a positive interaction the most widely used approach has been the idea that equal status, common goals, collaboration, and structural support during the interaction form the optimal conditions for a positive contact effect \citep[][]{Allport1954b, Pettigrew1969}. However, despite their prominence in guiding research on this topic, meeting the contact conditions does not seem to be necessary to finding positive effects of intergroup contact \citep[][]{Pettigrew2006}. Following Allport's original conditions, several additional conditions of optimal contact were proposed, which have largely received the same criticism of not capturing the underlying psychological mechanisms of why an interaction is perceived as positive \citep[for a critical discussion see][]{Pettigrew1986}.

@@

Beyond the well-being and psychosocial functioning perspective, need fulfillment has recently also seen application as a psychological mechanism in the intergroup relations literature. Social identity theory has famously focused on the role of self-esteem needs in understanding how people navigate intergroup contexts \citep[e.g.,][]{abrams1988}. A more recent example is the study of conflict and reconciliation where addressing differential needs of victims and perpetrators (i.e., the need for power and the need for morality respectively) increased willingness to reconcile \citep[][]{Shnabel2008}. And similarly, addressing a relevant need for identity continuity among refugees in Turkey bolstered resilience in the face of discrimination experiences \citep[][]{Celebi2017}. Most recently, an international cross-sectional study has applied a need-based approach to the effects of positive intergroup contact. The study found that minority members who generally perceive their positive intergroup contacts to be more empowering tended to report a smaller reduction in social change support than those who did not perceive their average interactions as empowering, especially when the inequality between groups was perceived as illegitimate \citep[][]{Hassler2021}. In short, an increasing amount of literature is emphasizing the significance of need satisfaction in comprehending intergroup relations.

It is thus not surprising that in their perspectives paper on intergroup contact, Dovidio and colleagues propose that: "To achieve truly constructive intergroup relations, it is important that intergroup exchanges meet the psychological needs of both majority- and minority-group members." \citep[][p. 6]{Dovidio2017}. A call that has thus far remained unanswered when it comes to the basic tenet that positive interactions drive the constructive part of intergroup contact.

@@

The psychological experience of need fulfillment is, thus, distinct from the content of the need (i.e., the motive or goal). The content could, for example, include physical motives (such as safety or hygiene) but also psychosocial motives \citep[such as acceptance or competence; e.g., see][]{pittman2007}. A need arises when a motive is thwarted or situationally active and relevant \citep[][]{leander2020, lewin1926e, gollwitzer1985e}. It is this perceived need and the perceived fulfillment of needs that we focus on in this article. This is not to say that considering specific motives is irrelevant to contact situations but instead we propose that the psychological experience of perceived need fulfillment is a core mechanism in understanding interaction quality perceptions, well-being, and outgroup attitudes.

To test such a proposal we can rely on adaptive and responsive survey designs that allow a tailored approach based on the participants' inputs \citep[e.g.,][]{Tourangeau2017}. In particular, we propose to ask the participants to report their main goal during the interaction in a short open-ended question (i.e., name the situationally relevant need content), and with reference to their own response, the participants can then indicate how much this need was fulfilled during the interaction (i.e., need fulfillment mechanism). Such an adaptive approach allows us to take the initial step of testing whether situational need fulfillment indeed generally predicts perceived interaction quality, well-being, and positive outgroup attitudes independent of need content.

\section{Intergroup Contact in Daily Life}
While we have argued that a need fulfillment mechanism is relevant to intergroup contact generally, its flexible and broad applicability might be ideally suited to address the pressing issue of understanding natural intergroup contacts outside the lab. Investigations of such `real-life interactions' often suffer from the difficulty that past intergroup contact research has either focused on the mechanisms of individual interactions in artificial lab studies (sometimes referred to as the intergroup interaction literature) or has focused on longer-term recall self-reports of natural interactions \citep[commonly referred to as the intergroup contact literature; also see][]{Pettigrew2006}. \citet[][]{MacInnis2015} have even pointed out that these two approaches tend to find conflicting effects --- where individual interactions (in the lab) have more negative effects and recall of real-world contact patterns have more positive effects for intergroup relations. We, thus, miss data that follows people in their diverse daily interactions and investigates the psychological mechanisms of contacts, especially as they compound over time. Even with extended intervention studies, the most fine-grained data available is usually limited to pre-post-control designs.

@@

In this appendix, we present the two additional robustness analyses that we conducted within each of the three studies. These analyses are specifically designed to check for alternative need models and to check that the effect of situation core need fulfillment is contact-specific. (1) To check for the role of alternate psychological needs, we add the fulfillment of self-determination theory needs (i.e., competence, autonomy, and relatedness) to the multi-level regression. We then also compare the model with models that predicts outgroup attitudes from self-determination theory need fulfillments or core need fulfillments only. (2) To ensure that the core need fulfillment is outgroup contact specific, we return to the full sample of intensive longitudinal measurements within each study and test whether there is an interaction effect of outgroup contact (vs. no outgroup contact) and core need fulfillment. We expect that the effect of core need fulfillment is specific to outgroup interactions and not merely due to a more need-fulfilled life in general.

@@

\subsection{Stability and Conclusion}
As with the main analyses, we use a forest plot and the associated meta-analytic estimates to assess the stability of the results across the three studies. The forest plot shows that the core need fulfillment is also consistently outgroup contact specific (\fgrref[-A]{fig:Robustness}) and remains a meaningful predictor of outgroup attitudes even when controlling for traditional psychological need fulfillments (\fgrref[-B]{fig:Robustness}). In sum, we thus find that across all three studies, psychological need fulfillment remained a robust and flexible predictor of positive outgroup attitudes.

@@

\begin{figure}
  \caption{Robustness Analyses}
  \label{fig:Robustness}
  \centering\includegraphics[width=\textwidth]{Figures/forestParametricRobustnessComb.png}
  %\begin{subfigure}{\textwidth}
  %  \caption{}
  %  \centering\includegraphics[width=0.65\textwidth]{Figures/forestParametricFERobustContact.png}
  %\end{subfigure}
  %\begin{subfigure}{\textwidth}
  %  \caption{}
  %  \centering\includegraphics[width=0.65\textwidth]{Figures/forestParametricFERobustSDT.png}
  %\end{subfigure}
  \caption*{Note: \\
  (a) Need Fulfillment and Intergroup Contact predicting Outgroup Attitudes (full sample).\\
  (b) Core Need Fulfillment predicting Outgroup Attitudes while controlling for self-determination theory needs (intergroup contact sample).\\
  General: Random effects meta-analytic results are presented for completeness.}
\end{figure}

>>>>

% Aim / Solution v.02: look at need fulfillment as a mechanism in daily interactions of migrants
% Proposal Paragraph:
We propose that one key to understanding how an interaction is perceived is to examine the level of need fulfillment it provided to an individual. As an example, if someone seeks acceptance by their interaction partner, and this need is fulfilled during the interaction, the person should rate the interaction and the group of the interaction partner more favorably. To test this idea, we collected three sets of real-life data from recent immigrants, assessing their daily interactions with majority group members, tracking situational needs, interaction quality, well-being, and outgroup attitudes.

\section{Need Mechanism in Intergroup Contact}
% Motivational mechanism:
% interaction quality is important
% past research has either focused on conditions or cognitive-affective processes
% neither are necessary nor explain why an interaction is positive
% Several meta-analytic reviews have found that positive intergroup contact reduces prejudice and increases positive attitudes in experimental and cross-sectional studies \citep[][]{Tropp2005, Pettigrew2006}, as well as in intergroup contact interventions outside the lab \citep[][]{Beelmann2014, Lemmer2015}. 
When it comes to what constitutes a positive interaction the most widely used approach has been the idea that equal status, common goals, collaboration, and structural support during the interaction form the optimal conditions for a positive contact effect \citep[][]{Allport1954b, Pettigrew1969}. However, despite their prominence in guiding research on this topic, meeting the contact conditions does not seem to be necessary to finding positive effects of intergroup contact \citep[][]{Pettigrew2006}. Following Allport's original conditions, several additional conditions of optimal contact were proposed, which have largely received the same criticism of not capturing the underlying psychological mechanisms of why an interaction is perceived as positive \citep[for a critical discussion see][]{Pettigrew1986}.

@@

Beyond the well-being and psychosocial functioning perspective, need fulfillment has recently also seen application as a psychological mechanism in the intergroup relations literature. Social identity theory has famously focused on the role of self-esteem needs in understanding how people navigate intergroup contexts \citep[e.g.,][]{abrams1988}. In the study of conflict and reconciliation, addressing differential needs of victims and perpetrators (i.e., the need for power and the need for morality respectively) increased willingness to reconcile \citep[][]{Shnabel2008}. And similarly, addressing a relevant need for identity continuity among refugees in Turkey bolstered resilience in the face of discrimination experiences \citep[][]{Celebi2017}. Most recently, an international cross-sectional study has applied a need-based approach to the effects of positive intergroup contact. The study found that minority members who generally perceive their positive intergroup contacts to be more empowering tended to report a smaller reduction in social change support than those who did not perceive their average interactions as empowering, especially when the inequality between groups was perceived as illegitimate \citep[][]{Hassler2021}. In short, an increasing amount of literature is emphasizing the significance of need satisfaction in comprehending intergroup relations.

It is thus not surprising that Dovidio and colleagues propose that: "To achieve truly constructive intergroup relations, it is important that intergroup exchanges meet the psychological needs of both majority- and minority-group members." \citep[][p. 6]{Dovidio2017}. A call that has thus far remained unanswered when it comes to the basic tenet that positive interactions drive the constructive part of intergroup contact.

@@

The psychological experience of need fulfillment is, thus, distinct from the content of the need (i.e., the motive or goal). The content could, for example, include physical motives (such as safety or hygiene) but also psychosocial motives \citep[such as acceptance or competence; e.g., see][]{pittman2007}. The experience of needing is a more general process that arises when any important motive is thwarted or situationally active and relevant \citep[][]{leander2020, lewin1926e, gollwitzer1985e}. It is this perceived need and the perceived fulfillment of needs that we focus on in this article. This is not to say that considering specific motives is irrelevant to contact situations but instead we propose that the psychological experience of perceived need fulfillment is a core mechanism in understanding interaction quality perceptions, well-being, and outgroup attitudes.

To test such a proposal, we can rely on adaptive and responsive survey designs that allow a tailored approach based on the participants' inputs \citep[e.g.,][]{Tourangeau2017}. In particular, we propose to ask the participants to report their main goal during the interaction in a short open-ended question (i.e., name the situationally relevant need content), and with reference to their own response, the participants can then indicate how much this need was fulfilled during the interaction (i.e., need fulfillment mechanism). Such an adaptive approach allows us to take the initial step of testing whether situational need fulfillment indeed generally predicts perceived interaction quality, well-being, and positive outgroup attitudes independent of need content.

\section{Intergroup Contact in Daily Life}
While we have argued that a need fulfillment mechanism is relevant to intergroup contact generally, its flexible and broad applicability might be ideally suited to address the pressing issue of understanding natural intergroup contacts outside the lab. Investigations of such `real-life interactions' often suffer from the difficulty that past intergroup contact research has either focused on the mechanisms of individual interactions in artificial lab studies (sometimes referred to as the intergroup interaction literature) or has focused on longer-term recall self-reports of natural interactions \citep[commonly referred to as the intergroup contact literature; also see][]{Pettigrew2006}. \citet[][]{MacInnis2015} have even pointed out that these two approaches tend to find conflicting effects --- where individual interactions (in the lab) have more negative effects and recall of real-world contact patterns have more positive effects for intergroup relations. We, thus, miss data following people in their diverse daily interactions and investigates the psychological mechanisms of contacts, especially as they compound over time. Even with extended intervention studies, the most fine-grained data available is usually limited to pre-post-control designs.

@@

In this appendix, we present the empirical details of our additional robustness analyses. These analyses are specifically designed to check for alternative models and contextualize our results. We (1) check whether core need fulfillment is indeed outgroup contact specific. For this analysis, we return to the full sample of intensive longitudinal measurements and test whether there is an interaction effect of outgroup contact (vs. no outgroup contact) and core need fulfillment. We expect that the effect of core need fulfillment is specific to outgroup interactions and not merely due to a more need-fulfilled life in general. We then check (2) whether the need mechanism is relevant to both planned and accidental outgroup interactions, and (3) extends to the more individual-focused experience of well-being. In a final set of analyses, we (4) check whether the need fulfillment mechanism is relevant to different types of need content (i.e., motives) and (5) remains relevant even when accounting for the fulfillment of fundamental psychological needs (i.e., self-determination theory needs: competence, autonomy, and relatedness).

@@

\subsection{Conclusion}
Across the wide variety of robustness analyses, we thus find that the experience of psychological need fulfillment is a robust and flexible predictor of positive outgroup attitudes even when accounting for a range of other and even alternate predictors. However, we also find that the core need fulfillment mechanism does not account for all need-related variance in outgroup attitudes. Notably, the fulfillment of relatedness (and to some extent competence) needs explained additional variance in outgroup attitudes. Nonetheless, the situational core need fulfillment remained extremely reliably a core predictor of outgroup attitudes.

@@

% \begin{figure}
%   \caption{Robustness Analyses}
%   \label{fig:Robustness}
%   \centering\includegraphics[width=\textwidth]{Figures/forestParametricRobustnessComb.png}
%   %\begin{subfigure}{\textwidth}
%   %  \caption{}
%   %  \centering\includegraphics[width=0.65\textwidth]{Figures/forestParametricFERobustContact.png}
%   %\end{subfigure}
%   %\begin{subfigure}{\textwidth}
%   %  \caption{}
%   %  \centering\includegraphics[width=0.65\textwidth]{Figures/forestParametricFERobustSDT.png}
%   %\end{subfigure}
%   \caption*{Note: \\
%   (a) Need Fulfillment and Intergroup Contact predicting Outgroup Attitudes (full sample).\\
%   (b) Core Need Fulfillment predicting Outgroup Attitudes while controlling for self-determination theory needs (intergroup contact sample).\\
%   General: Random effects meta-analytic results are presented for completeness.}
% \end{figure}
